---
title: "Getting started with ocean3d"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ocean3d}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 6,
  fig.height = 4
)
```

ocean3d analyses marine habitat in three dimensions: area *and* depth. It
represents species ranges, fisheries and oceanographic variables as stacked
`terra` rasters, so they can be overlapped, masked and summarised by depth.

# Extracting Voxel Values Within a Species' 3D Range

This article walks the core operation the package exists for: take a species
range **polygon**, turn it into a 3D domain, and read the values a
depth-stratified environmental raster carries inside that domain.

Three objects and two conversions do the whole job:

| Object | What it is |
|---|---|
| `SpatEnvelope` | The species' 3D range. Two layers, `depth_min` and `depth_max`; **depth is the cell value**, so the vertical interval varies per cell. |
| `SpatVoxel` | The environmental raster. One layer per standard depth, named `{variable}_depth={value}`; **depth is the layer index**. |
| occupancy voxel | The envelope re-expressed on the raster's depth axis, so the two can be combined layer by layer. |

Everything here is synthetic and runs — no downloads, no local data files. The
grids are small enough to print, so you can check every step against the
numbers. The [articles](#next-steps) listed at the end apply the same workflow
to real data.

```{r setup}
library(ocean3d)
```

## Step 1: A study grid and a seafloor

> **Example data.** Steps 1 and 2 build synthetic stand-ins for the inputs a
> real analysis downloads. In practice the seafloor comes from GEBCO
> (`load_gebco_bathymetry()`) and the environmental raster from the World Ocean
> Atlas (`woa_download()`, `woa_load_nc()`).

The study grid defines the geometry everything else is aligned to. The seafloor
is depth in **positive metres increasing downward**, the package's convention
throughout — GEBCO-style elevation has to be flipped before it gets here.

```{r grid}
grid <- terra::rast(
  nrows = 20, ncols = 20,
  xmin = 0, xmax = 20, ymin = 0, ymax = 20,
  crs = "EPSG:4326"
)

# A shelf that deepens from north to south, 20 m down to 900 m.
seafloor <- terra::rast(grid)
terra::values(seafloor) <- rep(seq(20, 900, length.out = 20), each = 20)

terra::plot(seafloor, main = "Seafloor depth (m)")
```

## Step 2: A WOA-shaped environmental raster, as a `SpatVoxel`

World Ocean Atlas products arrive as one layer per standard depth. Here that
shape is built by hand, as example data: temperature falling with depth, on a
latitudinal gradient. `as_voxel()` attaches the layer names and validates the
depth axis. With real WOA data, `woa_load_nc()` reads the depth layers and
`as_voxel()` wraps them in the same way.

```{r voxel}
standard_depths <- c(0, 50, 100, 200, 500)

lat <- terra::init(terra::rast(grid), "y")
t_layers <- lapply(standard_depths, function(d) 26 - 0.03 * d - 0.4 * lat)

t_annual <- as_voxel(
  terra::rast(t_layers),
  depths = standard_depths,
  varname = "t_an"
)

names(t_annual)
```

`depths()` reads that axis back out. It is the accessor to reach for whenever
you need the numeric depths rather than the layer names:

```{r depths}
depths(t_annual)
```

## Step 3: The species range polygon, as a `SpatEnvelope`

`vect_to_envelope()` rasterises the polygon onto the study grid and attaches
the depth limits, giving the species' 3D domain. With real data the polygon is
a species range map, e.g. from the IUCN Red List, and the depth limits come
from its assessment (`fetch_species_assessments()`).

The seafloor is not a special argument — it is simply one more `depth_max`
constraint. Per cell the **shallowest** `depth_max` wins, so a species with a
nominal 600 m limit is clamped to the bed wherever the bed is shallower.

```{r envelope}
poly <- terra::vect(
  "POLYGON ((4 2, 16 2, 16 18, 4 18, 4 2))",
  crs = "EPSG:4326"
)

range_env <- vect_to_envelope(
  polygon   = poly,
  template  = grid,
  depth_min = 0,
  depth_max = list(600, seafloor)
)

range_env
```

Depth is the cell value here, and the interval is per cell — the northern rows
are clamped to a shallow bed, the southern rows reach the full 600 m:

```{r envelope-values}
range(terra::values(range_env[["depth_max"]]), na.rm = TRUE)
```

```{r envelope-plot, fig.height = 3.5}
terra::plot(range_env)
```

Cells where the bed sits above `depth_min` have no water column left and drop
out entirely. A species living below 150 m cannot occupy the northernmost row
of the polygon, where the bed is only about 113 m deep:

```{r envelope-deep}
deep_env <- vect_to_envelope(
  polygon   = poly,
  template  = grid,
  depth_min = 150,
  depth_max = list(600, seafloor)
)

# Occupied cells: the full polygon, then the deeper-living species.
sum(!is.na(terra::values(range_env[["depth_max"]])))
sum(!is.na(terra::values(deep_env[["depth_max"]])))
```

```{r envelope-deep-plot}
terra::plot(deep_env[["depth_max"]], main = "depth_max, depth_min = 150 m")
terra::lines(poly)
```

## Step 4: Put the envelope on the voxel's depth axis

The envelope and the voxel store depth in different roles, so they cannot be
combined directly. `envelope_to_voxel()` re-expresses the envelope on the
voxel's depth levels, giving a `1`/`NA` occupancy voxel with one layer per
depth.

Each depth level stands for a slab of water, not a single depth. WOA levels
stand for the water halfway to their neighbours, so `bounds = "midpoint"` is
the convention to use with WOA data. See
`vignette("depth-levels", package = "ocean3d")` for how slabs are placed and
why it matters.

```{r occupancy}
occ <- envelope_to_voxel(
  range_env,
  depths = depths(t_annual),
  bounds = "midpoint"
)

# Cells occupied at each standard depth.
terra::global(occ, "sum", na.rm = TRUE)
```

```{r occupancy-plot, fig.height = 5}
terra::plot(
  occ, nc = 3, legend = FALSE, col = "steelblue",
  main = paste(depths(occ), "m")
)
```

## Step 5: Read the values inside the range

With both objects on the same depth axis, masking is ordinary `terra`:

```{r mask}
in_range <- terra::mask(t_annual, occ)

names(in_range)
```

The layer names survive, so a single depth can still be pulled out by name —
sea surface temperature within the range, for instance:

```{r surface}
sst <- in_range[["t_an_depth=0"]]
summary(terra::values(sst, na.rm = TRUE)[, 1])
```

```{r surface-plot}
terra::plot(sst, main = "Surface temperature within the range")
terra::lines(poly)
```

### Vertical profile

`depths()` also accepts a character vector, which is what makes a per-layer
`terra::global()` result easy to turn into a profile — the layer names are on
the rows, not on a raster:

```{r profile}
per_depth <- terra::global(in_range, c("mean", "min", "max"), na.rm = TRUE)
per_depth$depth <- depths(rownames(per_depth))
per_depth$n_cells <- terra::global(!is.na(in_range), "sum", na.rm = TRUE)[, 1]

per_depth[, c("depth", "n_cells", "mean", "min", "max")]
```

The cell count falls with depth: the range is clamped to the bed, so the
northern rows drop out of the deeper levels.

```{r profile-plot, fig.width = 4, fig.height = 4}
plot(
  per_depth$mean, per_depth$depth,
  type = "b", ylim = rev(range(per_depth$depth)),
  xlab = "Mean temperature (°C)", ylab = "Depth (m)"
)
```

### Summary statistics across the whole 3D range

Summaries over the masked voxel are a few lines of base R over
`terra::values()`, which returns a cells × depths matrix. `n_cells` counts
occupied **voxels** (cell × depth), while `n_surface_cells` counts the map
cells with data at any depth:

```{r summarise}
summarise_voxel <- function(v, name) {
  vals <- terra::values(v)
  out <- c(
    min             = suppressWarnings(min(vals, na.rm = TRUE)),
    max             = suppressWarnings(max(vals, na.rm = TRUE)),
    mean            = suppressWarnings(mean(vals, na.rm = TRUE)),
    n_surface_cells = sum(rowSums(!is.na(vals)) > 0),
    n_cells         = sum(!is.na(vals)),
    n_depths        = terra::nlyr(v)
  )
  # An all-NA range gives -Inf / Inf from min() and max().
  out[is.infinite(out)] <- NA_real_
  names(out) <- paste(name, names(out), sep = "_")
  as.data.frame(as.list(out))
}

summarise_voxel(in_range, "temperature")
```

Across several variables, and several species, this is an `lapply()` over the
same three lines. Steps 4 and 5 need not be done by hand: given the envelope
directly, `terra::mask()` places it on each raster's own depth levels before it
masks.

```{r summarise-many}
# Example data: a second variable, on its own set of standard depths.
do_annual <- as_voxel(
  terra::rast(lapply(c(0, 100, 300), function(d) 300 - 0.2 * d + 2 * lat)),
  depths = c(0, 100, 300),
  varname = "o_an"
)

env_rasters <- list(temperature = t_annual, oxygen = do_annual)

do.call(cbind, unname(Map(function(v, nm) {
  summarise_voxel(terra::mask(v, range_env, bounds = "midpoint"), nm)
}, env_rasters, names(env_rasters))))
```

Each environmental raster gets its own occupancy voxel. Note `oxygen` reports
three depths against temperature's five: the levels are read from the raster
being masked, never assumed shared.

## Step 6: Collapsing back to an envelope

`voxel_to_envelope()` is the return trip: it reduces a voxel to the envelope
bounding it. A predicate decides what counts as present, so it can bound a
subset of the values rather than just the data extent — here, the depths over
which the range is warmer than 20 °C:

```{r collapse}
warm <- voxel_to_envelope(in_range, fun = function(x) x > 20)
range(terra::values(warm[["depth_max"]]), na.rm = TRUE)
```

This conversion is **lossy**, so a round trip through `envelope_to_voxel()`
does not give back the voxel you started from:

- An envelope holds one continuous interval per cell, so any interior gap in
  the voxel is filled in — a cell satisfying the predicate at 0 m and 200 m but
  not at 100 m still comes back as `[0, 200]`.
- The envelope records depth *levels*, not the slabs they stand for. A cell
  warm only at the 0 m level comes back as `[0, 0]`, although that level stands
  for 0–25 m of water. Expanded again, such a zero-thickness interval occupies
  no slab at all, and the cell is lost.

Where either matters, stay in the voxel.

## Summary

Once the inputs are loaded, the whole workflow is two calls:

```{r recap, eval = FALSE}
# 1. the species' 3D range
range_env <- vect_to_envelope(poly, grid, depth_min = 0,
                              depth_max = list(600, seafloor))

# 2. the values inside the range
in_range <- terra::mask(t_annual, range_env, bounds = "midpoint")
```

The second is the one line to reach for whenever a voxel needs restricting to
a species' per-cell depth window. `terra::mask()` is depth-aware for the
package's 3D classes: it places the envelope on the voxel's own depth levels,
as `envelope_to_voxel()` did by hand in Step 4, and masks layer by layer.

To restrict a voxel to a polygon and a single depth band across the whole area
instead — no per-cell window — use `extract_to_area()`.

## Next steps

These articles apply the workflow to real data. They download data or need API
keys, so they are on the package website rather than installed with the
package:

- [Extracting 3D Environmental Data for a Single Species](https://marine-biodiversity-conservation-lab.github.io/ocean3d/articles/woa-environmental-extraction-single-species.html)
  — an IUCN range and World Ocean Atlas 2023 climatologies.
- [Depth-Stratified Fishing Effort from Global Fishing Watch](https://marine-biodiversity-conservation-lab.github.io/ocean3d/articles/gfw-fishing-effort-3d.html)
  — Global Fishing Watch effort by gear type, overlapped with a species range.
- [3D Volume Overlap Between Species and Fisheries](https://marine-biodiversity-conservation-lab.github.io/ocean3d/articles/bangladesh-fisheries-3d-overlap.html)
  — `calc_volume_overlap()` on species ranges and fishery footprints.
