---
title: "Getting started with ocean3d"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ocean3d}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 6,
  fig.height = 4
)
```

ocean3d analyses marine habitat in three dimensions: area *and* depth. It
can represent things like species ranges, fisheries and oceanographic variables in 3D, 
so they can be overlapped, masked and summarised by depth. ocean3d does this by representing 
space with a stacked raster representation, which will be explained further in the next section. 

# Extracting 3D Values Within a Species' 3D Range

This article walks through one core operation of the package: taking a species
range **polygon**, convert it to 3D, and read the values from a 3D (raster stack by depth) 
oceanographic dataset. 

| Object | What it is |
|---|---|
| `SpatEnvelope` | The species' 3D range. Two layers, `depth_min` and `depth_max`; **depth is the cell value**, so the vertical interval varies per cell. |
| `SpatVoxel` | The environmental raster. One layer per standard depth, named `{variable}_depth={value}`; **depth is the layer index**. |

In this Getting Started vignette, we use small synthetic representative datasets. Therefore, 
this requires no downloads and no local data files, with grids that are small enough to print, 
so you can follow every step. The [articles](#next-steps) listed at the end apply the same workflow
to real data.

```{r setup}
library(ocean3d)
```

## Step 1: A study grid and a seafloor

> **Example data.** Steps 1 and 2 build synthetic stand-ins for the inputs a
> real analysis downloads. In practice the seafloor comes from GEBCO
> (`load_gebco_bathymetry()`) and the environmental raster from the World Ocean
> Atlas (`woa_download()`, `woa_load_nc()`).

The study grid defines the geometry everything else is aligned to. The seafloor
is depth in **positive metres increasing downward**, which is the package's convention. 
GEBCO's elevation values have to be flipped. 

```{r grid}
grid <- terra::rast(
  nrows = 20, ncols = 20,
  xmin = 0, xmax = 20, ymin = 0, ymax = 20,
  crs = "EPSG:4326"
)

# A shelf that deepens from north to south, 20 m down to 900 m.
seafloor <- terra::rast(grid)
terra::values(seafloor) <- rep(seq(20, 900, length.out = 20), each = 20)

terra::plot(seafloor, main = "Seafloor depth (m)")
```

## Step 2: A WOA-shaped environmental raster, as a `SpatVoxel`

World Ocean Atlas data products are organized as one layer per standard depth. 
Here, this shape is built by hand as example data. This synthetic dataset shows
temperature decreasing with depth, on a latitudinal gradient. `as_voxel()` 
attaches the layer names and validates the depth axis. With real WOA data, 
`woa_load_nc()` reads in the depth layers.

```{r voxel}
standard_depths <- c(0, 50, 100, 200, 500)

lat <- terra::init(terra::rast(grid), "y")
t_layers <- lapply(standard_depths, function(d) 26 - 0.03 * d - 0.4 * lat)

# This works with the woa_load_nc() return as input as well
t_annual <- as_voxel(
  terra::rast(t_layers),
  depths = standard_depths,
  varname = "t_an"
)

names(t_annual)
```

`depths()` reads that axis back out. This is the helper function that 
returns the numeric depths rather than the layer names:

```{r depths}
depths(t_annual)
```

## Step 3: The species range polygon, as a `SpatEnvelope`

`vect_to_envelope()` rasterises the polygon onto the study grid and attaches
the depth limits, giving the species' 3D domain. For real data, the polygon is
a species range map, e.g. from the IUCN Red List Spatial Download, with the 
depth limits coming from its IUCN Red List assessment 
(`fetch_species_assessments()`).

The seafloor is not a special argument, and is treated as simply one more `depth_max`
constraint. Per cell the **shallowest** `depth_max` wins, so a species with a 600 m 
limit is clamped to the seabed, in any cells where the seabed is shallower than the 
species `depth_max`. 

```{r envelope}
poly <- terra::vect(
  "POLYGON ((4 2, 16 2, 16 18, 4 18, 4 2))",
  crs = "EPSG:4326"
)

range_env <- vect_to_envelope(
  polygon   = poly,
  template  = grid,
  depth_min = 0,
  depth_max = list(600, seafloor)
)

range_env
```

Depth is the cell value here, and the interval is per cell — the northern rows
are clamped to a shallow bed, the southern rows reach the full 600 m:

```{r envelope-values}
range(terra::values(range_env[["depth_max"]]), na.rm = TRUE)
```

```{r envelope-plot, fig.height = 3.5}
terra::plot(range_env)
```

Cells where the bed sits above `depth_min` have no water column left and drop
out entirely. A species living below 150 m cannot occupy the northernmost row
of the polygon, where the bed is only about 113 m deep:

```{r envelope-deep}
deep_env <- vect_to_envelope(
  polygon   = poly,
  template  = grid,
  depth_min = 150,
  depth_max = list(600, seafloor)
)

# Occupied cells: the full polygon, then the deeper-living species.
sum(!is.na(terra::values(range_env[["depth_max"]])))
sum(!is.na(terra::values(deep_env[["depth_max"]])))
```

```{r envelope-deep-plot}
terra::plot(deep_env[["depth_max"]], main = "depth_max, depth_min = 150 m")
terra::lines(poly)
```

## Step 4: Put the envelope on the voxel's depth axis

The envelope and the voxel store depth in different roles, so they cannot be
combined directly. `envelope_to_voxel()` re-expresses the envelope on the
voxel's depth levels, giving a `1`/`NA` occupancy voxel with one layer per
depth.

Each depth level stands for a slab of water, not a single depth. WOA levels
stand for the water halfway to their neighbours, so `bounds = "midpoint"` is
the convention to use with WOA data. See
`vignette("depth-levels", package = "ocean3d")` for how slabs are placed and
why it matters.

```{r occupancy}
occ <- envelope_to_voxel(
  range_env,
  depths = depths(t_annual),
  bounds = "midpoint"
)

# Cells occupied at each standard depth.
terra::global(occ, "sum", na.rm = TRUE)
```

```{r occupancy-plot, fig.height = 5}
terra::plot(
  occ, nc = 3, legend = FALSE, col = "steelblue",
  main = paste(depths(occ), "m")
)
```

## Step 5: Read the values inside the range

With both objects on the same depth axis, masking can use `terra::mask()`:

```{r mask}
in_range <- terra::mask(t_annual, occ)

names(in_range)
```

The layer names are maintained, so that a single depth can still be pulled out by name. 
For example, sea surface temperature within the range:

```{r surface}
sst <- in_range[["t_an_depth=0"]]
summary(terra::values(sst, na.rm = TRUE)[, 1])
```

```{r surface-plot}
terra::plot(sst, main = "Surface temperature within the range")
terra::lines(poly)
```

### Vertical profile

`depths()` also accepts a character vector, making a per-layer
`terra::global()` result easy to turn into a profile. 

```{r profile}
per_depth <- terra::global(in_range, c("mean", "min", "max"), na.rm = TRUE)
per_depth$depth <- depths(rownames(per_depth))
per_depth$n_cells <- terra::global(!is.na(in_range), "sum", na.rm = TRUE)[, 1]

per_depth[, c("depth", "n_cells", "mean", "min", "max")]
```

The cell count falls with depth: the range is clamped to the bed, so the
northern rows drop out of the deeper levels.

```{r profile-plot, fig.width = 4, fig.height = 4}
plot(
  per_depth$mean, per_depth$depth,
  type = "b", ylim = rev(range(per_depth$depth)),
  xlab = "Mean temperature (°C)", ylab = "Depth (m)"
)
```

## Step 6: Collapsing back to an envelope

`voxel_to_envelope()` is the return trip: it reduces a voxel to the envelope
bounding it. A predicate decides what counts as present, so it can bound a
subset of the values rather than just the data extent — here, the depths over
which the range is warmer than 20 °C:

```{r collapse}
warm <- voxel_to_envelope(in_range, fun = function(x) x > 20)
range(terra::values(warm[["depth_max"]]), na.rm = TRUE)
```

This conversion is **lossy**, so a round trip through `envelope_to_voxel()`
does not give back the voxel you started from:

- An envelope holds one continuous interval per cell, so any interior gap in
  the voxel is filled in — a cell satisfying the predicate at 0 m and 200 m but
  not at 100 m still comes back as `[0, 200]`.
- The envelope records depth *levels*, not the slabs they stand for. A cell
  warm only at the 0 m level comes back as `[0, 0]`, although that level stands
  for 0–25 m of water. Expanded again, such a zero-thickness interval occupies
  no slab at all, and the cell is lost.

Where either matters, stay in the voxel.

## Summary

Once the inputs are loaded, the whole workflow is two calls:

```{r recap, eval = FALSE}
# 1. the species' 3D range
range_env <- vect_to_envelope(poly, grid, depth_min = 0,
                              depth_max = list(600, seafloor))

# 2. the values inside the range
in_range <- terra::mask(t_annual, range_env, bounds = "midpoint")
```

The second is the one line to reach for whenever a voxel needs restricting to
a species' per-cell depth window. `terra::mask()` is depth-aware for the
package's 3D classes: it places the envelope on the voxel's own depth levels,
as `envelope_to_voxel()` did by hand in Step 4, and masks layer by layer.

To restrict a voxel to a polygon and a single depth band across the whole area
instead — no per-cell window — use `extract_to_area()`.

## Next steps

These articles apply the workflow to real data. They download data or need API
keys, so they are on the package website rather than installed with the
package:

- [Extracting 3D Environmental Data for a Single Species](https://marine-biodiversity-conservation-lab.github.io/ocean3d/articles/woa-environmental-extraction-single-species.html)
  — an IUCN range and World Ocean Atlas 2023 climatologies.
- [Depth-Stratified Fishing Effort from Global Fishing Watch](https://marine-biodiversity-conservation-lab.github.io/ocean3d/articles/gfw-fishing-effort-3d.html)
  — Global Fishing Watch effort by gear type, overlapped with a species range.
- [3D Volume Overlap Between Species and Fisheries](https://marine-biodiversity-conservation-lab.github.io/ocean3d/articles/bangladesh-fisheries-3d-overlap.html)
  — `calc_volume_overlap()` on species ranges and fishery footprints.

## Feedback and contributing

ocean3d is new and under active development, so feedback from first-time users
is especially valuable. If something in this vignette was unclear, a function
did not behave as you expected, or there is functionality you wish existed,
please open an issue on
[GitHub Issues](https://github.com/Marine-Biodiversity-Conservation-Lab/ocean3d/issues).
A small reproducible example makes bugs much quicker to fix.

Contributions are welcome, from fixing a typo to adding a new workflow. If you
have written code for 3D marine analyses that others could reuse, adding it to
ocean3d broadens its impact. See the
[contributing guide](https://github.com/Marine-Biodiversity-Conservation-Lab/ocean3d/blob/main/CONTRIBUTING.md)
for how to get set up, and the
[good first issues](https://github.com/Marine-Biodiversity-Conservation-Lab/ocean3d/labels/good%20first%20issue)
for places to start. Please note that ocean3d is released with a
[Contributor Code of Conduct](https://github.com/Marine-Biodiversity-Conservation-Lab/ocean3d/blob/main/CODE_OF_CONDUCT.md);
by contributing, you agree to abide by its terms.
